MSTK: Mesh Toolkit, v 1.7 - DRAFT

### 2010-03-01 docs/MSTK.tex
Added section about handling physics variables and boundary conditions

Even though MSTK is written in C, MSTK is designed in an object
oriented manner. Mesh entities, mesh attributes and entity lists are
considered to be objects which have their private data and a set of
operators to query and manipulate this data. Even though methods may
be devised to circumvent the data hiding in MSTK, this is highly
discouraged. If you think that a particular operator is not efficient
or you don'thave access to some mesh data, please contact the
developer for assistance.

@@

\subsection{Getting meshes in and out of MSTK}

@@

>>>>

Even though MSTK is written in C, MSTK is designed in an object
oriented manner. Mesh entities, mesh attributes and entity lists are
considered to be like objects which have their private data and a set
of operators to query and manipulate this data. Even though methods
may be devised to circumvent the data hiding in MSTK, since it is
written in C, this is highly discouraged. If you think that a
particular operator is not efficient or you don'thave access to some
mesh data, please contact the MSTK developer for assistance.

@@

@@

\section{Getting meshes in and out of MSTK}

\section{Using MSTK in a numerical simulation code}

In addition to mesh generation and mesh manipulation algorithms, MSTK
can be used to write mesh-based numerical simulation codes such as
Finite element, finite volume and finite difference methods. MSTK is
particularly suited to developing methods which operate on general
unstructured meshes. There is some overhead to using operators to
access mesh data instead of directly accessing arrays; however, we
believe that the ease of managing complex mesh data through an API
outweighs this cost. Without a framework like MSTK, applications have
to devise their own unstructured mesh data structures and the code is
littered with direct accesses to these data structures (often
complicated array-based data structures). This makes even small
changes to the data structure very onerous because thousands to tens
of thousands of lines of code are affected. In the end, individual
developers create adhoc patches (often duplicated in different ways)
to overcome any deficiencies in the data structures and eventually the
code becomes unmanageable. We believe this is poor software
engineering.

\subsection{Simulation Variables}

When using MSTK in a physics based code, the question of storing
physics variables (node velocities, element
pressures) is an important one.

There are two ways of storing physics in combination with MSTK. The
first is to use the attributes feature in MSTK. Scalar variables can
be stored with entities as double precision values of mesh attributes;
vector and tensor variables can be stored as pointer variables. When
storing vectors as pointers, it is more efficient to allocate a large
array to store the vectors for all entities and store pointers into
that large array with the individual entities. This way the
application does not allocate a huge number of small arrays. This
method is very convenient as the variables are readily available using
the handle of the entity, i.e., by calling {\bf MEnt\_Get\_AttVal}.
However, this method could be somewhat inefficient and also requires a
paradigm shift in how codes are written.

The other way of managing physics variables is to directly allocate
linear arrays for each variable and using the entity ID to reference
entries in this array. For example, one might have a linear array {\bf
  rho} for the densities of all elements and we use {\bf
  rho[MR\_ID(r)-1]} to find the density of a particular element {\bf
  r}. The ID of elements is obtained quite easily and stays static
unless {\bf MESH\_Renumber} is called explicitly. If there is mesh
modification, there will be intermediate parts of this array which
will no longer be relevant but this is easily managed. This method is
closer to traditional approaches and can be adopted quite easily. It
could be more efficient than using attributes but is less
``object-oriented.''

\subsection{Boundary Conditions, Material Properties}

Applying boundary conditions in an MSTK-based simulation code can be
much easier than in a ``traditional'' simulation code, {\em if} the
mesh carries the right classification information. In a
``traditional'' simulation code, boundary conditions are specified
using one of two ways:

\begin{itemize}
\item An implicit geometric entity is defined and the boundary
  condition is specified on that geometry, e.g., normal velocity is 1.0
  for all nodes on the plane [1.35 2.7695 1]. This requires one to check
  the coordinates of all nodes to see if they lie on this plane and
  apply the boundary condition on those nodes.

\item Entity sets like node sets or face sets are defined as a
  collection of entity IDs on which a particular boundary conditions
  are applied. The user has pre-selected a set of nodes and applied
  the boundary condition one them.
\end{itemize}

MSTK, on the other hand, provides for an easier way to specify
boundary conditions, initial conditions and material properties
provided the mesh has classification information. For example, if we
knew that model face 51 has a pressure boundary condition on it, we
only need to find all faces that are classified on model face 51
(MF\_GEntDim(f)=2, MF\_GEntID(f)=51) and specify the pressure on that
face in the code. The input file for the boundary conditions does not
have to explicitly list any face sets nor do we have to use error
prone geometric checks on the entire mesh to apply boundary
conditions. In fact, if the mesh is directly tied to a geometric model
(using a GEntity\_ptr), it is possible to specify a boundary condition
on a complex surface such as a Trimmed NURBS surface and still apply
the condition without error-prone geometric checks or large input
files. Similarly, material properties can be specified per geometric
model region and applied automatically to every mesh region classified
in that model region without the need for element sets in input files.
